\LARGE \bf Mean Estimation from Adaptive One-bit Measurements

### 2017-10-02 Writeups/onebit_adaptive_Allerton_V3.tex
minor edit

Thm.~\ref{thm:opt_one_step} suggests the following adaptive encoding and estimation scheme: 
\begin{itemize}
\item Initialization: set $P_0(t) = \pi(\theta)$.
\item For $n\geq 1$:
\begin{enumerate}
\item Update the prior as
\begin{align}
P_n(t) = & P(\theta=t |M^n) \\
& = \frac{ P\left( \theta=t | M^{n-1} \right) P(M_n | \theta = t , M^{n-1})  } { P(M_n | M^{n-1} )} \nonumber \\ 
& = \alpha_n  P_{n-1}(t) \Phi\left(M_n \frac{ t - \tau_{n-1} }{\sigma} \right), \label{eq:density_update}
\end{align}
where $\alpha_n$ is a normalization coefficient that equals to
\[
\alpha_n = \left(\int_{\mathbb R} P_{n-1}(t) \Phi\left(M_n \frac{t- \tau_{n-1} }{\sigma} \right)  dt \right)^{-1}. 
\]
\item The $n$th estimate for $\theta$ is the conditional expectation of $\theta$ given $M^n$, namely
\begin{equation}
\theta_n = \mathbb E \left[ \theta| M^n\right] = \int_{-\infty}^\infty t P_n(t) dt. \label{eq:estimator_update}
\end{equation}
\item Solve equation \eqref{eq:fixed_point} with the updated prior $P_n(t)$ instead of $P(d\theta)$. Note that since the standard normal cdf $\Phi(x)$ is log-concave, the updated prior $P_n(t)$ remains log-concave and thus a unique solution to \eqref{eq:fixed_point} is guaranteed by Lem.~\ref{lem:unique}. 
\item Update the $(n+1)$th message as
\begin{equation}\label{eq:message_update}
M_{n+1} = \sgn(X_{n+1}-\tau_n)
\end{equation}
\end{enumerate}
\end{itemize}
Since equation \eqref{eq:fixed_point} has no analytic solution in general, it is hard to derive the asymptotic behavior of the estimator defined by \eqref{eq:estimator_update} and \eqref{eq:message_update}. We conjecture, however, that it attains the asymptotic relative efficiency of $\sigma^2\pi/2$, as can be observed from the numerical simulation illustrated in Fig.~\ref{fig:adaptive_error}. Also shown in Fig.~\ref{fig:adaptive_error} are the normalized MSE of the asymptotically optimal estimator defined by \eqref{eq:sgd_alg} and \eqref{eq:sgd_est}, as well as the MSE achieved by the sample mean for the same sample realization.

\begin{figure}
\begin{center}
\begin{tikzpicture}
\node at (0,0) {\includegraphics[scale=0.4]{one_bit_adpative}};
%\node[color = blue,  minimum width=1cm, align = left] at (1.5,1.5) {assymptotically optimal \eqref{eq:sgd_est}};
%\node[color = red, minimum width=1cm, align = left] at (1.5,0.55) {one-step optimal \eqref{eq:message_update}} ;
\node[rotate = 90, scale = 0.7] at (-3.8,0) {$n \mathbb E \left(\widehat{\theta}_n - \theta \right)^2$};
\node[scale = 0.7] at (0,-2.4) {$n$};
%\draw[dashed] (-3.5,1.18) node[left, scale = 0.7] {$\frac{\pi}{2}$}-- (-3,1.18);
\end{tikzpicture}
\caption{Normalized empirical risk $n\left(\widehat{\theta}_n-\theta\right)^2$ versus number of samples $n$ for $500$ Monte Carlo trials. In each trial, $\theta$ is chosen uniformly in the interval $(-3,3)$. \label{fig:adaptive_error}  }
\end{center}
\end{figure}

@@

\subsubsection*{Proof of Lem.~\ref{lem:bound_intervals}}
We use induction on $n \in \mathbb N$. For the base case $n=1$ we have 
\begin{equation} \label{eq:induction_base}
\frac{  \phi^2(x)} 
{\Phi(x) \left(1 - \Phi(x) \right) },
\end{equation}
which is the weight function in the probit analysis and is known to be a strictly decreasing function of $|x|$ \cite{Samford1953}. In particular, the maximum of \eqref{eq:induction_base} is obtained at $x=0$ where it equals $2/\pi$. \par
Assume now that \eqref{eq:bound_intervals} holds for all integers up to some $n = N-1$ and consider the case $n = N$. The maximal value of \eqref{eq:bound_intervals} is attained for the same $(x_1,\ldots,x_N) \in \mathbb R^N$ that attains the maximal value of 
\begin{align*}
& g(x_1,\ldots, x_N) \triangleq 2 \log \left(  \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k) \right) - \\
&  \log
\left( \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right)
-\log \left(1 -  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right) \\
& = 2 \log \delta_N - \log \Delta_N - \log \left(1 - \Delta_N  \right),
\end{align*}
where we denoted $\delta_N \triangleq \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k)$ and $\Delta_N =  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k)$. The derivative of $g(x_1,\ldots,x_N)$ with respect to $x_k$ is given by
\[
\frac{\partial  g}{\partial x_k} = \frac{2 (-1)^{k+1} \phi'(x_k)}{\delta_N} -\frac{(-1)^{k+1} \phi(x_k)}{\Delta_N } + \frac{(-1)^{k+1} \phi(x_k)}{1-\Delta_N }.
\]
Using the fact that $\phi'(x) = -x \phi(x)$, we conclude that the gradient of $g$ vanishes only if
\[
x_k = \frac{\delta_N}{2} \left( \frac{1}{\Delta_N} - \frac{1}{1-\Delta_N} \right),\quad k=1,\ldots,N.
\]
In particular, the condition above implies $x_1 = \ldots = x_N$. If $N$ is odd then for $x_1=\ldots =x_N$ we have that the LHS of \eqref{eq:bound_intervals} equals
\[
\frac{\phi(x_1)^2}{ \Phi(x_1) (1 - \Phi(x_1))},
\]
which was shown to be smaller than $\pi/2$. If $N$ is even, then for any constant $c$ the limit of the LHS of \eqref{eq:bound_intervals} as $(x_1,\ldots,x_N)\rightarrow (c,\ldots,c)$ exists and equals zero. Therefore, the maximum of the LHS of \eqref{eq:bound_intervals} is not attained at the line $x_1=\ldots=x_N)$. We now consider the possibility that the LHS of \eqref{eq:bound_intervals} is maximized at the borders, as one or more of the coordinates of $(x_1,\ldots,x_N)$ approaches plus or minus infinity. For simplicity we only consider the cases where $x_N$ goes to minus infinity or $x_1$ goes to plus infinity (the general case where the first $m$ coordinates goes to infinity or the last $m$ to minus infinity is obtained using similar arguments). Assume first $x_N \rightarrow -\infty$. Then the LHS of \eqref{eq:bound_intervals} equals
\begin{align*}
\frac{ \left(  \sum_{k=1}^{N-1} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k) \right)\left(1- \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k)  \right) } ,
\end{align*}
which is smaller than $2/\pi$ by the induction hypothesis. Assume now that $x_1 \rightarrow \infty$. Then the LHS of
\eqref{eq:bound_intervals} equals
\begin{align*}
& \frac{ \left(  \sum_{k=2}^{N} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( 1 + \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k) \right)\left(1- 1 - \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k)  \right) }  \\
& = \frac{ \left(  -\sum_{m=1}^{N} (-1)^{m+1}\phi(x'_m) \right)^2} 
{\left( 1 - \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m}) \right)\left( \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m})  \right) },
\end{align*}
where $x'_{m} = x_{m+1}$. The last expression is also smaller than $2/\pi$ by the induction hypothesis. This proves Lem.~\ref{lem:bound_intervals}. \\

>>>>

Thm.~\ref{thm:opt_one_step} suggests the following adaptive encoding and estimation scheme: 
\begin{itemize}
\item Initialization: set $P_0(t) = \pi(\theta)$.
\item For $n\geq 1$:
\begin{enumerate}
\item Update the prior as
\begin{align}
P_n(t) = & P(\theta=t |M^n) \\
& = \frac{ P\left( \theta=t | M^{n-1} \right) P(M_n | \theta = t , M^{n-1})  } { P(M_n | M^{n-1} )} \nonumber \\ 
& = \alpha_n  P_{n-1}(t) \Phi\left(M_n \frac{ t - \tau_{n-1} }{\sigma} \right), \label{eq:density_update}
\end{align}
where $\alpha_n$ is a normalization coefficient that equals to
\[
\alpha_n = \left(\int_{\mathbb R} P_{n-1}(t) \Phi\left(M_n \frac{t- \tau_{n-1} }{\sigma} \right)  dt \right)^{-1}. 
\]
\item The $n$th estimate for $\theta$ is the conditional expectation of $\theta$ given $M^n$, namely
\begin{equation}
\theta_n = \mathbb E \left[ \theta| M^n\right] = \int_{-\infty}^\infty t P_n(t) dt. \label{eq:estimator_update}
\end{equation}
\item Solve equation \eqref{eq:fixed_point} with the updated prior $P_n(t)$ instead of $\pi(d\theta)$. Note that since the standard normal cdf $\Phi(x)$ is log-concave, the updated prior $P_n(t)$ remains log-concave and thus a unique solution to \eqref{eq:fixed_point} is guaranteed by Lem.~\ref{lem:unique}. 
\item Update the $(n+1)$th message as
\begin{equation}\label{eq:message_update}
M_{n+1} = \sgn(X_{n+1}-\tau_n)
\end{equation}
\end{enumerate}
\end{itemize}
Since equation \eqref{eq:fixed_point} has no analytic solution, it is hard to derive the asymptotic behavior of the estimator defined by \eqref{eq:estimator_update} and \eqref{eq:message_update}. We conjecture that it attains the asymptotic relative efficiency of $\sigma^2\pi/2$ as can be observed from the numerical simulation illustrated in Fig.~\ref{fig:adaptive_error}. Also shown in Fig.~\ref{fig:adaptive_error} are the normalized MSE of the asymptotically optimal estimator defined by \eqref{eq:sgd_alg} and \eqref{eq:sgd_est}, as well as the MSE achieved by the sample mean for the same sample realization.

\begin{figure}
\begin{center}
\begin{tikzpicture}
\node at (0,0) {\includegraphics[scale=0.4]{one_bit_adaptive}};
%\node[color = blue,  minimum width=1cm, align = left] at (1.5,1.5) {assymptotically optimal \eqref{eq:sgd_est}};
%\node[color = red, minimum width=1cm, align = left] at (1.5,0.55) {one-step optimal \eqref{eq:message_update}} ;
\node[rotate = 90, scale = 0.7] at (-3.8,0) {$n \mathbb E \left(\widehat{\theta}_n - \theta \right)^2$};
\node[scale = 0.7] at (0,-2.4) {$n$};
%\draw[dashed] (-3.5,1.18) node[left, scale = 0.7] {$\frac{\pi}{2}$}-- (-3,1.18);
\end{tikzpicture}
\caption{Normalized empirical risk $n\left(\widehat{\theta}_n-\theta\right)^2$ versus number of samples $n$ for $500$ Monte Carlo trials. In each trial, $\theta$ is chosen uniformly in the interval $(-3,3)$. \label{fig:adaptive_error}  }
\end{center}
\end{figure}

@@

\subsubsection*{Proof of Lem.~\ref{lem:bound_intervals}}
We use induction on $n \in \mathbb N$. For the base case $n=1$ we have 
\begin{equation} \label{eq:induction_base}
\frac{  \phi^2(x)} 
{\Phi(x) \left(1 - \Phi(x) \right) },
\end{equation}
which is the weight function in the probit analysis and is known to be a strictly decreasing function of $|x|$ \cite{Samford1953}. In particular, the maximum of \eqref{eq:induction_base} is obtained at $x=0$ where it equals $2/\pi$. \par
Assume now that \eqref{eq:bound_intervals} holds for all integers up to some $n = N-1$ and consider the case $n = N$. The maximal value of \eqref{eq:bound_intervals} is attained for the same $(x_1,\ldots,x_N) \in \mathbb R^N$ that attains the maximal value of 
\begin{align*}
& g(x_1,\ldots, x_N) \triangleq 2 \log \left(  \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k) \right) - \\
&  \log
\left( \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right)
-\log \left(1 -  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k) \right) \\
& = 2 \log \delta_N - \log \Delta_N - \log \left(1 - \Delta_N  \right),
\end{align*}
where we denoted $\delta_N \triangleq \sum_{k=1}^{N} (-1)^{k+1} \phi(x_k)$ and $\Delta_N =  \sum_{k=1}^N (-1)^{k+1} \Phi(x_k)$. The derivative of $g(x_1,\ldots,x_N)$ with respect to $x_k$ is given by
\[
\frac{\partial  g}{\partial x_k} = \frac{2 (-1)^{k+1} \phi'(x_k)}{\delta_N} -\frac{(-1)^{k+1} \phi(x_k)}{\Delta_N } + \frac{(-1)^{k+1} \phi(x_k)}{1-\Delta_N }.
\]
Using the fact that $\phi'(x) = -x \phi(x)$, we conclude that the gradient of $g$ vanishes only if
\[
x_k = \frac{\delta_N}{2} \left( \frac{1}{\Delta_N} - \frac{1}{1-\Delta_N} \right),\quad k=1,\ldots,N.
\]
In particular, the condition above implies $x_1 = \ldots = x_N$. If $N$ is odd then for $x_1=\ldots =x_N$ we have that the LHS of \eqref{eq:bound_intervals} equals
\[
\frac{\phi(x_1)^2}{ \Phi(x_1) (1 - \Phi(x_1))},
\]
which was shown to be not larger than $2/pi$. If $N$ is even, then for any constant $c$ the limit of the LHS of \eqref{eq:bound_intervals} as $(x_1,\ldots,x_N)\rightarrow (c,\ldots,c)$ exists and equals zero. Therefore, the maximum of the LHS of \eqref{eq:bound_intervals} is not attained at the line $x_1=\ldots=x_N)$. We now consider the possibility that the LHS of \eqref{eq:bound_intervals} is maximized at the borders, as one or more of the coordinates of $(x_1,\ldots,x_N)$ approaches plus or minus infinity. For simplicity we only consider the cases where $x_N$ goes to minus infinity or $x_1$ goes to plus infinity (the general case where the first $m$ coordinates goes to infinity or the last $m$ to minus infinity is obtained using similar arguments). Assume first $x_N \rightarrow -\infty$. Then the LHS of \eqref{eq:bound_intervals} equals
\begin{align*}
\frac{ \left(  \sum_{k=1}^{N-1} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k) \right)\left(1- \sum_{k=1}^{N-1} (-1)^{k+1} \Phi(x_k)  \right) } ,
\end{align*}
which is smaller than $2/\pi$ by the induction hypothesis. Assume now that $x_1 \rightarrow \infty$. Then the LHS of
\eqref{eq:bound_intervals} equals
\begin{align*}
& \frac{ \left(  \sum_{k=2}^{N} (-1)^{k+1}\phi(x_k) \right)^2} 
{\left( 1 + \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k) \right)\left(1- 1 - \sum_{k=2}^{N} (-1)^{k+1} \Phi(x_k)  \right) }  \\
& = \frac{ \left(  -\sum_{m=1}^{N} (-1)^{m+1}\phi(x'_m) \right)^2} 
{\left( 1 - \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m}) \right)\left( \sum_{m=1}^{N-1} (-1)^{m+1} \Phi(x'_{m})  \right) },
\end{align*}
where $x'_{m} = x_{m+1}$. The last expression is also smaller than $2/\pi$ by the induction hypothesis. This proves Lem.~\ref{lem:bound_intervals}. \\
